---
title: "Analysis of Paired End Rna Seq data  "
author: "Aashi Sharma"
date: '2023-01-06'
output :html_document:code_folding: "hide"  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#set working directory

```{r,message=FALSE,warning=FALSE}
setwd("C:/Users/Aashi/Desktop/DGE")
```

##loading library

```{r,message=FALSE,warning=FALSE}
# load the libraries    
library(DESeq2)
library(apeglm)
library(ggplot2)
library(pheatmap)
library(cowplot)
library(hrbrthemes)
library(viridis)
library(plotly)
library(dplyr)
library(tidyverse)
library(org.Hs.eg.db)
library(fgsea)
library("msigdbr")
```

###DGE

```{r,message=FALSE,warning=FALSE}
# DGE
##loading count matrix
gse = read.csv("GSE92945_Fibroblast_RNAseq_counts_1.csv", header = TRUE, row.names = 1)
info = read.table("ind.txt", header = T,sep ='\t') 

##creating a data matrix  
dds = DESeqDataSetFromMatrix(round(gse), info, ~conditions)

##removing lowly expressed genes
keep = rowSums(counts(dds)) >= 10
dds = dds[keep,]

##main deseq
ddsDE = DESeq(dds)

##export normalized read counts
normCOunts = counts(ddsDE, normalized = T)
write.csv(normCOunts,"normalised_data.csv")

##DESEQ results
res = results(ddsDE, alpha = 0.05)

##output DESeq result
resOrdered = res[order(res$padj),]
write.csv(resOrdered, "resOrdered1.csv")


  
```

#plotting

```{r}
# loading and preparing data for plotting 
##pca and disp(plots)
vsat = vst(ddsDE, blind = FALSE)

pca = plotPCA(vsat, intgroup = "conditions")
disp = plotDispEsts(ddsDE)

ggplotly(pca)


##load_the_deseq2_results_data_and_metadata  

normCOunt = read.csv('normalised_data.csv', row.names = 1)

deSeqRes = read.csv('resOrdered.csv',row.names = 1)

deSeqRes$sig =ifelse(deSeqRes$padj <= 0.05, 'yes', 'no')

deSeqRes = na.omit(deSeqRes)
```

##ggplots

```{r,message=FALSE}
# ggplots
##ma_plot 
maplot = ggplot(deSeqRes, aes(x = log10(baseMean), y = log2FoldChange, color = sig)) + geom_point(alpha = 0.7)+scale_size(range = c(1.4, 19), name = "MA PLOT")+ scale_color_viridis(discrete = TRUE, guide=FALSE)+theme_ipsum()+theme(legend.position = "none")
  
 

##volcano plot 
volcano = ggplot(deSeqRes, aes(x = log2FoldChange, y = -log10(padj), color = sig)) + geom_point(alpha = 0.7)+scale_size(range = c(1.9, 20), name = "MA PLOT")+ scale_color_viridis(discrete = TRUE, guide=FALSE)+theme_ipsum()+theme(legend.position = "none")
volcano = volcano + ylim(0,15)



##pheatmap
signi = subset(deSeqRes, padj <= 0.05)
allsig= merge(normCOunt, signi, by = 0)

sigcounts = allsig[,2:11]
row.names(sigcounts) = allsig$Row.names 
```
```{r,warning=FALSE}
pheat = pheatmap(log2(sigcounts + 1), scale = 'row', show_rownames = F, treeheight_row = 0, treeheight_col = 0)

###cowplot
plot_grid(volcano,pheat[[4]])
plot_grid(maplot,volcano)
###interactive plot using ggplotly 
ggplotly(maplot, tooltip = "text")
ggplotly(volcano)

```


```

#GSEA

```{r,message=FALSE }
# GSEA
 
rest <- read_csv("resOrdered1.csv")
head(rest)
rest = rest %>%
  rename(gene = ...1 )
##creating a mapping table
library(org.Hs.eg.db)
ens2symbol = AnnotationDbi::select(org.Hs.eg.db,keys = rest$gene, 
                                   columns="SYMBOL",
                                   keytype="ENSEMBL")
ens2symbol <- as_tibble(ens2symbol)
view(ens2symbol)

rest <- inner_join(rest, ens2symbol, by=c("gene"="ENSEMBL"))
rest

res2 <- rest %>% 
  dplyr::select(SYMBOL, stat) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(SYMBOL) %>% 
  summarize(stat=mean(stat))
res2

ranks <- deframe(res2)
head(ranks, 20)

####loading reference enrichment database 

Hs.GOBP <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")
Hs.Reactome <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")
Hs.Hallmark <- msigdbr(species = "Homo sapiens", category = "H")

Hs.GOBP.Entrez <- Hs.GOBP %>% split(x = .$entrez_gene, f = .$gs_name)
Hs.Hallmark.Entrez <- Hs.Hallmark %>% split(x = .$entrez_gene, f = .$gs_name)
Hs.GOBP.Symbol <- Hs.GOBP %>% split(x = .$gene_symbol, f = .$gs_name)
Hs.Hallmark.Symbol <- Hs.Hallmark %>% split(x = .$gene_symbol, f = .$gs_name)



#####Now that we have everything lets run fgsea on it  
fgseaRes.Hallmark <- fgsea::fgseaMultilevel(pathways=Hs.Hallmark.Symbol,stats = ranks, eps =0)
```

GSEA but prettier

```{r}
# Make the Results tidy
fgseaResTidy <- fgseaRes.Hallmark %>%
  as_tibble() %>%
  arrange(desc(NES))

fgseaResTidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj) %>% 
  DT::datatable()

head(fgseaResTidy,10)

##plotting_the_results 
gea_plot =  ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA") + 
  theme_minimal()
ggplotly(gea_plot)  

```

#playing around with different pathways

```{r,warning=FALSE}
# trying_out_in_different_pathways
##kegg
kegg = fgsea(pathways=gmtPathways("c2.cp.kegg.v6.2.symbols.gmt"), ranks) %>% 
  as_tibble() %>% 
  arrange(padj)%>%
  DT::datatable()

##go
go = fgsea(pathways=gmtPathways("c5.all.v6.2.symbols.gmt"), ranks) %>%
  as_tibble()%>%
  arrange(padj)%>%
  DT::datatable()

##miR

miR = fgsea(pathways=gmtPathways("c3.mir.v6.2.symbols.gmt"), ranks) %>%
  as_tibble()%>%
  arrange(padj)%>%
  DT::datatable()

sessionInfo()
```
